Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2025-04-11 manuscript_latex/manuscript.tex
Add appendix stuff.

Then the log-likelihood function of the location- and scale- invariant Box-Cox power transformation is defined as:
\begin{equation}
\label{eqn:box-cox-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{BC}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
\end{definition}

@@

Then the log-likelihood function of the location- and scale- invariant Yeo-Johnson power transformation is defined as:
\begin{equation}
\label{eqn:yeo-johnson-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{YJ}}^{\lambda, x_0, s} = & -\frac{n}{2} \log\left[2 \pi \sigma^2\right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{YJ}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \mathop{\mathrm{sgn}}(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
\end{definition}

@@

\begin{equation}
\label{eqn:box-cox-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{BC}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
\end{definition}

@@

\begin{equation}
\label{eqn:yeo-johnson-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{YJ}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \mathop{\mathrm{sgn}}(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
\end{definition}

@@

We used simulated data to assess invariance to location and scale of the
proposed power transformations, to develop the empirical central
normality test, and to determine weighting for robust transformations.
The \(\lambda\) parameter for conventional power transformations (Eqn.
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well
as \(\lambda\), \(x_0\) and \(s\) parameters for location- and
scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant}
and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA
algorithm for derivative-free bound constraint optimisation (Powell
2009) through maximum likelihood estimation. The required algorithms
were implemented in the \texttt{power.transform} R software package
(Zwanenburg and Löck 2024b) (version 1.0.1). Of note, the
\texttt{power.transform} package shifts feature values into the positive
domain if negative or zero values are present for Box-Cox power
transformations.

@@

Location- and scale-invariant power transformations are intended to
yield improved transformations to normality in the presence of large
shifts in location, distributions that due to location and scale are not
centered near zero, or both. Earlier, we assessed these transformations
using simulated data. In the following, they are evaluated using
examples from real datasets. We focus on the Yeo-Johnson transformation
because of its ability to handle features with negative values. Results
for Box-Cox transformations are shown in Appendix D.

@@

We used 231 datasets containing at least one numeric feature from the
Penn Machine Learning Benchmarks collection (Romano et al. 2022). In
this collection, 114 datasets correspond to regression tasks and 117
datasets to classification tasks. Using the familiar auto-machine
learning library (Zwanenburg and Löck 2024a) (version 1.5.0), each
dataset was used to train a model for each of 32 process configurations.
Each process configuration specifies the learner (generalised linear
model, L1-regularised linear models (Lasso), gradient boosted linear
model, or random forest), transformation method (none, conventional
Yeo-Johnson, robust invariant Yeo-Johnson, robust invariant Yeo-Johnson
with empirical central normality test (rejecting transformations with
\(p \leq 0.01\)), and normalisation method (none,
\(z\)-standardisation), yielding 32 distinct configurations. Before each
experiment, each dataset was randomly split into a training (70\%) and
holdout test (30\%) set five times. Thus, a total of 36960 models were
created. Each model was then evaluated using the holdout test set using
one of two metrics, i.e.~the root relative squared error (RRSE) for
regression tasks and the area under the receiver operating
characteristic curve (AUC) for classification tasks.

@@

\section{Log-likelihood functions for location and scale
invariant power transformation}\label{appendix-a-log-likelihood-functions-for-location-and-scale-invariant-power-transformation}

Location and scale-invariant Box-Cox and Yeo-Johnson transformations are
parametrised using location \(x_0\) and scale \(s\) parameters, in
addition to transformation parameter \(\lambda\). This leads to the
following transformations. The location and scale-invariant Box-Cox
transformation is:

@@

\subsection{Location- and scale-invariant Box-Cox
transformation}\label{location--and-scale-invariant-box-cox-transformation}

@@

Similarly to \citet{Raymaekers2024-zf}, sample weights \(w_i\) are
introduced to facilitate robust power transformations. The weighted
log-likelihood of the location- and scale-invariant Box-Cox
transformation is:

@@

\subsection{Location- and scale-invariant Yeo-Johnson
transformation}\label{location--and-scale-invariant-yeo-johnson-transformation}

@@

\section{Optimisation of transformation parameters}\label{appendix-b-optimisation-of-transformation-parameters}

@@

Below we will focus on criteria based on the empirical density function
and those based on skewness and kurtosis of the transformed featured.
Other potential criteria, such as the Shapiro-Wilk test statistic
\citep{Shapiro1965-zd} are not investigated here. In the case of the
Shapiro-Wilk test statistic this is because of lack of scalability to
features with many (\(> 5000\)) instances, and because adapting the test
statistic to include weights is not straightforward.

@@

The first class of criteria is based on the empirical distribution
function (EDF). Transformation parameters are then fit through
minimisation of the distance between the empirical distribution function
\(F_{\epsilon}\) and the cumulative density function (CDF) of the normal
distribution \(F_{\mathcal{N}}\). Let
\(F_{\epsilon}\left(x_i \right) = \frac{i - 1/3}{n + 1/3}\) be the
empirical probability of instance \(i\). The normal distribution is
parametrised by location parameter \(\mu\) and scale parameter
\(\sigma\), both of which have to be estimated from the data. For
non-robust power transformations, \(\mu\) and \(\sigma\) are sample mean
and sample standard deviation, respectively. For robust power
transformations, we estimate \(\mu\) and \(\sigma\) as Huber M-estimates
of location and scale of the transformed feature
\(\phi^{\lambda, x_0, s} (\mathbf{X})\) \citep{Huber1981-su}.

@@

Here \(w_i\) are weights, and \(\mu\) and \(\sigma\) are location and
scale parameters. For non-robust power transformations, all \(w_i = 1\).
Note that this criterion is not the same as the Anderson-Darling test
statistic \citep{Anderson1952-gz}, which involves solving (or
approximating) an integral function, contains an extra scalar
multiplication term, and does not include weights. The Anderson-Darling
criterion seeks to minimise the squared Euclidean distance between the
EDF and the normal CDF, with differences at the upper and lower end of
the normal CDF receiving more weight than those at the the centre of the
CDF.

\subsubsection{Cram{\'e}r-von Mises
criterion}\label{cramer-von-mises-criterion}

The Cram{\'e}r-von Mises criterion is also based on the empirical
distribution function of \(\mathbf{X}\). We define the Cram{\'e}r-von Mises
criterion as follows:

@@

Here \(w_i\) are weights, and \(\mu\) and \(\sigma\) are location and
scale parameters. For non-robust power transformations, all \(w_i = 1\).
The criterion is similar to the Cram{\'e}r-von Mises test statistic \citep{Cramer1928-rc, Von_Mises1928-ef},
aside from a additive scalar value and the introduction of
weights. This criterion, like the Anderson-Darling criterion, seeks to
minimise the squared Euclidean distance between the EDF and the normal
CDF. Unlike the Anderson-Darling criterion, this criterion weights all
instances equally.

For conventional power transformations with a fixed shift parameter, the
transformation \(\phi^{\lambda, x_0, s} (\mathbf{X})\) may be
substituted by \(\phi^{\lambda} (\mathbf{X})\) in the definition of the
Cram{\'e}r-von Mises criterion.

@@

The D'Agostino criterion defined here follows the D'Agostino \(K^2\)
test statistic \citep{DAgostino1990-kp}. This test statistic is
composed of two separate test statistics, one of which is related to
skewness, and the other to kurtosis. Both test statistics are computed
in several steps. Let us first define \(\nu=\sum_{i=1}^n w_i\). Thus for
non-robust power transformations, \(\nu = n\).

For the skewness test statistic we first compute \citep{DAgostino1990-kp}:

@@

For the kurtosis test statistic we first compute \citep{Anscombe1983-nz, DAgostino1990-kp}:

@@

The main difference between the above criterion and the Jarque-Bera test
statistic \citep{Jarque1980-hw} is that a scalar multiplication is absent.

@@

\section{Simulations with other optimisation criteria}\label{appendix-c-simulations-with-other-optimisation-criteria}

Invariance of location- and scale-invariant power transformations was
assessed using the optimisation criteria in 
\ref{appendix-b-optimisation-of-transformation-parameters}. This follows the simulation in section \ref{invariance-to-location-and-scale}, where MLE was
used for optimization. In short, we first randomly drew \(10000\) values
from a normal distribution:
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)\),
or equivalently
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)\).
The second distribution was a right-skewed normal distribution
\(\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)\).
The third distribution was a left-skewed normal distribution
\(\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)\).
We then computed transformation parameter \(\lambda\) using the original
definitions (equations \ref{eqn:box-cox-original} and
\ref{eqn:yeo-johnson-original}) and the location- and scale-invariant
definitions (equations \ref{eqn:box-cox-invariant} and
\ref{eqn:yeo-johnson-invariant}) for each distribution using different
optimisation criteria. To assess location invariance, a positive value
\(d_{\text{shift}}\) was added to each distribution with
\(d_{\text{shift}} \in [1, 10^6]\). Similarly, to assess scale
invariance, each distribution was multiplied by a positive value
\(d_{\text{scale}}\), where \(d_{\text{scale}} \in [1, 10^6]\).

@@

{\centering \includegraphics[width=1\linewidth]{figure_appendix_1}

@@

\section{Experimental results using location- and scale-invariant Box-Cox transformation}
\label{appendix-d-experimental-results-using-location--and-scale-invariant-box-cox-transformation}

The effect of using location- and scale-invariant transformations was
investigated using real-world datasets.

\subsection{Invariance}\label{sec-app:invariance}

Results for Box-Cox transformations of features without outliers are
shown in Figure \ref{fig:experimental-results-invariance-appendix}.

@@

{\centering \includegraphics[width=1\linewidth]{figure_appendix_2}

@@

\caption{Quantile-quantile plots for several datasets: age of patients with lung cancer (top row); penguin body mass (middle row); and latitude coordinates of houses sold in Ames, Iowa (bottom row). Multiple quantile-quantile plots are shown: for the original feature (left column); the feature transformed using the conventional Box-Cox transformation and Raymaekers and Rousseeuw's robust adaptation (middle column); and the feature transformed using the non-robust and robust location- and-scale invariant Box-Cox transformations (right column).}\label{fig:experimental-results-invariance-appendix}
\end{figure}

\subsubsection{Age of patients with lung
cancer}\label{sec-app:age-of-patients-with-lung-cancer}

Applying conventional and invariant Box-Cox transformations to age of
patients with lung cancer \citep{Loprinzi1994-cd} yielded the following
results: no transformation (sum of residuals with normal distribution
\(\sum r_i = 16.5\)); conventional transformation (\(\lambda = 1.9\),
\(\sum r_i = 11.5\), \(\mu_{BC} = 1.6 \cdot 10^3\),
\(\sigma_{BC} = 0.4 \cdot 10^3\)); Raymaekers and Rousseeuw's robust
adaptation (\(\lambda = 1.9\), \(\sum r_i = 11.5\),
\(\mu_{BC} = 1.6 \cdot 10^3\), \(\sigma_{BC} = 0.4 \cdot 10^3\));
location- and scale-invariant transformation (\(\lambda = 1.7\),
\(\sum r_i = 11.6\), \(\mu_{BC} = 1.9\), \(\sigma_{BC} = 0.8\)); and
robust location- and scale-invariant transformation (\(\lambda = 1.5\),
\(\sum r_i = 11.6\), \(\mu_{BC} = 3.6\), \(\sigma_{BC} = 1.2\)).

Compared to location- and scale-invariant Yeo-Johnson transformations,
the Box-Cox transformations do not reduce residuals compared to
conventional variants.

\subsubsection{Penguin body mass}\label{sec-app:penguin-body-mass}

Applying conventional and invariant Box-Cox transformations to the body
mass of penguins \citep{Gorman2014-eo} yielded the
following results: no transformation (residual sum \(\sum r_i = 48.0\));
conventional transformation (\(\lambda = -0.5\), \(\sum r_i = 32.2\),
\(\mu_{BC} = 2.1\), \(\sigma_{BC} = 4 \cdot 10^{-3}\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = -0.5\), \(\sum r_i = 32.2\),
\(\mu_{BC} = 2.1\), \(\sigma_{BC} = 4 \cdot 10^{-3}\)); location- and
scale-invariant transformation (\(\lambda = 0.5\), \(\sum r_i = 27.3\),
\(\mu_{BC} = 0.3\), \(\sigma_{BC} = 0.6\)); and robust location- and
scale-invariant transformation (\(\lambda = 0.2\), \(\sum r_i = 25.2\),
\(\mu_{BC} = 0.4\), \(\sigma_{BC} = 0.6\)).

Just as for location- and scale-invariant Yeo-Johnson transformations,
Box-Cox transformations produced a lower overall residual sum compared
to their conventional counterparts. Similarly, conventional
transformations led to low standard deviation \(\sigma_{YJ}\) of the
body mass feature after transformation.

\subsubsection{Latitude in the Ames housing
dataset}\label{sec-app:latitude-in-the-ames-housing-dataset}

>>>>

Then the log-likelihood function of the location- and scale- invariant Box-Cox power transformation is defined as:
\begin{equation}
\label{eqn:box-cox-invariant-log-likelihood}
\begin{split}
\mathcal{L}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{BC}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
\end{definition}

@@

Then the log-likelihood function of the location- and scale- invariant Yeo-Johnson power transformation is defined as:
\begin{equation}
\label{eqn:yeo-johnson-invariant-log-likelihood}
\begin{split}
\mathcal{L}_{\text{YJ}}^{\lambda, x_0, s} = & -\frac{n}{2} \log\left[2 \pi \sigma^2\right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{YJ}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \mathop{\mathrm{sgn}}(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
\end{definition}

@@

\begin{equation}
\label{eqn:box-cox-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{L}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{BC}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
\end{definition}

@@

\begin{equation}
\label{eqn:yeo-johnson-weighted-invariant-log-likelihood}
\begin{split}
\mathcal{L}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi_{\text{YJ}}^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \mathop{\mathrm{sgn}}(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
\end{definition}

@@

We used simulated data to assess invariance to location and scale of the
proposed power transformations, to develop the empirical central
normality test, and to determine weighting for robust transformations.
The \(\lambda\) parameter for conventional power transformations (Eqn.
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well
as \(\lambda\), \(x_0\) and \(s\) parameters for location- and
scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant}
and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA
algorithm for derivative-free bound constraint optimisation (Powell
2009) through maximum likelihood estimation. The required algorithms
were implemented in the \texttt{power.transform} R software package
(Zwanenburg and L\"ock 2024b) (version 1.0.1). Of note, the
\texttt{power.transform} package shifts feature values into the positive
domain if negative or zero values are present for Box-Cox power
transformations.

@@

Location- and scale-invariant power transformations are intended to
yield improved transformations to normality in the presence of large
shifts in location, distributions that due to location and scale are not
centered near zero, or both. Earlier, we assessed these transformations
using simulated data. In the following, they are evaluated using
examples from real datasets. We focus on the Yeo-Johnson transformation
because of its ability to handle features with negative values. Results
for Box-Cox transformations are shown in Appendix E.

@@

We used 231 datasets containing at least one numeric feature from the
Penn Machine Learning Benchmarks collection (Romano et al. 2022). In
this collection, 114 datasets correspond to regression tasks and 117
datasets to classification tasks. Using the familiar auto-machine
learning library (Zwanenburg and L\"ock 2024a) (version 1.5.0), each
dataset was used to train a model for each of 32 process configurations.
Each process configuration specifies the learner (generalised linear
model, L1-regularised linear models (Lasso), gradient boosted linear
model, or random forest), transformation method (none, conventional
Yeo-Johnson, robust invariant Yeo-Johnson, robust invariant Yeo-Johnson
with empirical central normality test (rejecting transformations with
\(p \leq 0.01\)), and normalisation method (none,
\(z\)-standardisation), yielding 32 distinct configurations. Before each
experiment, each dataset was randomly split into a training (70\%) and
holdout test (30\%) set five times. Thus, a total of 36960 models were
created. Each model was then evaluated using the holdout test set using
one of two metrics, i.e.~the root relative squared error (RRSE) for
regression tasks and the area under the receiver operating
characteristic curve (AUC) for classification tasks.

@@

\section{Appendix A: Log-likelihood functions for location- and
scale-invariant power
transformation}\label{appendix-a}

Location- and scale-invariant Box-Cox and Yeo-Johnson transformations
are parametrised using location \(x_0\) and scale \(s\) parameters, in
addition to transformation parameter \(\lambda\). This leads to the
following transformations. The location- and scale-invariant Box-Cox
transformation is:

@@

\subsection{Location- and scale-invariant Box-Cox
transformation}\label{location--and-scale-invariant-box-cox-transformation-appendix}

@@

Similarly to Raymaekers and Rousseeuw (2024), sample weights \(w_i\) are
introduced to facilitate robust power transformations. The weighted
log-likelihood of the location- and scale-invariant Box-Cox
transformation is:

@@

\subsection{Location- and scale-invariant Yeo-Johnson
transformation}\label{location--and-scale-invariant-yeo-johnson-transformation-appendix}

@@

\section{Appendix B: Optimisation of transformation
parameters}\label{appendix-b}

@@

Below we will focus on criteria based on the empirical density function
and those based on skewness and kurtosis of the transformed featured.
Other potential criteria, such as the Shapiro-Wilk test statistic
(Shapiro and Wilk 1965) are not investigated here. In the case of the
Shapiro-Wilk test statistic this is because of lack of scalability to
features with many (\(> 5000\)) instances, and because adapting the test
statistic to include weights is not straightforward.

@@

The first class of criteria is based on the empirical distribution
function (EDF). Transformation parameters are then fit through
minimisation of the distance between the empirical distribution function
\(F_{\epsilon}\) and the cumulative density function (CDF) of the normal
distribution \(F_{\mathcal{N}}\). Let
\(F_{\epsilon}\left(x_i \right) = \frac{i - 1/3}{n + 1/3}\) be the
empirical probability of instance \(i\). The normal distribution is
parametrised by location parameter \(\mu\) and scale parameter
\(\sigma\), both of which have to be estimated from the data. For
non-robust power transformations, \(\mu\) and \(\sigma\) are sample mean
and sample standard deviation, respectively. For robust power
transformations, we estimate \(\mu\) and \(\sigma\) as Huber M-estimates
of location and scale of the transformed feature
\(\phi^{\lambda, x_0, s} (\mathbf{X})\) (Huber 1981).

@@

Here \(w_i\) are weights, and \(\mu\) and \(\sigma\) are location and
scale parameters. For non-robust power transformations, all \(w_i = 1\).
Note that this criterion is not the same as the Anderson-Darling test
statistic (Anderson and Darling 1952), which involves solving (or
approximating) an integral function, contains an extra scalar
multiplication term, and does not include weights. The Anderson-Darling
criterion seeks to minimise the squared Euclidean distance between the
EDF and the normal CDF, with differences at the upper and lower end of
the normal CDF receiving more weight than those at the the centre of the
CDF.

\subsubsection{Cramér-von Mises
criterion}\label{cramuxe9r-von-mises-criterion}

The Cramér-von Mises criterion is also based on the empirical
distribution function of \(\mathbf{X}\). We define the Cramér-von Mises
criterion as follows:

@@

Here \(w_i\) are weights, and \(\mu\) and \(\sigma\) are location and
scale parameters. For non-robust power transformations, all \(w_i = 1\).
The criterion is similar to the Cramér-von Mises test statistic Mises
(1928), aside from a additive scalar value and the introduction of
weights. This criterion, like the Anderson-Darling criterion, seeks to
minimise the squared Euclidean distance between the EDF and the normal
CDF. Unlike the Anderson-Darling criterion, this criterion weights all
instances equally.

For conventional power transformations with a fixed shift parameter, the
transformation \(\phi^{\lambda, x_0, s} (\mathbf{X})\) may be
substituted by \(\phi^{\lambda} (\mathbf{X})\) in the definition of the
Cramér-von Mises criterion.

@@

The D'Agostino criterion defined here follows the D'Agostino \(K^2\)
test statistic (D'Agostino and Belanger 1990). This test statistic is
composed of two separate test statistics, one of which is related to
skewness, and the other to kurtosis. Both test statistics are computed
in several steps. Let us first define \(\nu=\sum_{i=1}^n w_i\). Thus for
non-robust power transformations, \(\nu = n\).

For the skewness test statistic we first compute (D'Agostino and
Belanger 1990):

@@

For the kurtosis test statistic we first compute (D'Agostino and
Belanger 1990; Anscombe and Glynn 1983):

@@

The main difference between the above criterion and the Jarque-Bera test
statistic (Jarque and Bera 1980) is that a scalar multiplication is
absent.

@@

\subsection{Simulations with other optimisation
criteria}\label{simulations-with-other-optimisation-criteria}

Invariance of location- and scale-invariant power transformations was
assessed using the optimisation criteria in Appendix B. 
This follows the simulation in the main manuscript, where MLE was
used for optimization. In short, we first randomly drew \(10000\) values
from a normal distribution:
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)\),
or equivalently
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)\).
The second distribution was a right-skewed normal distribution
\(\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)\).
The third distribution was a left-skewed normal distribution
\(\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)\).

We then computed transformation parameter \(\lambda\) using the original
definitions (equations \ref{eqn:box-cox-original} and
\ref{eqn:yeo-johnson-original}) and the location- and scale-invariant
definitions (equations \ref{eqn:box-cox-invariant} and
\ref{eqn:yeo-johnson-invariant}) for each distribution using different
optimisation criteria. To assess location invariance, a positive value
\(d_{\text{shift}}\) was added to each distribution with
\(d_{\text{shift}} \in [1, 10^6]\). Similarly, to assess scale
invariance, each distribution was multiplied by a positive value
\(d_{\text{scale}}\), where \(d_{\text{scale}} \in [1, 10^6]\).

@@

{\centering \includegraphics[width=1\linewidth]{appendix_figure_1}

@@

\section{Appendix C: Central normality test and empirical central
normality test critical
values}\label{appendix-c}

Critical values for the central normality test and its empirical variant
are found in Table \ref{tab:central-normality-critical-statistic} and
Table \ref{tab:empirical-central-normality-critical-statistic},
respectively.

\begin{table}
\begin{center}
\caption{Critical values of test statistic $\tau_{\alpha, n, \kappa = 0.80}$ for
\textbf{central normality} at $\kappa = 0.80$, as a function of significance 
level $\alpha$ and number of instances $n$. The values shown must be divided by $100$.}
\label{tab:central-normality-critical-statistic}
\begin{tabular}{l | r r r r r r r r r}

\toprule
$n$ \textbackslash $\alpha$ & 0.001 & 0.01 & 0.025 & 0.05 & 0.1 & 0.2 & 0.5 & 0.8 & 0.9 \\

\midrule
    5 & 210.79 & 95.75 & 66.27 & 49.69 & 35.93 & 27.60 & 19.29 & 13.41 & 11.00 \\
   10 &  48.04 & 31.44 & 26.96 & 23.91 & 21.11 & 18.16 & 13.66 & 10.14 &  8.67 \\
   20 &  29.13 & 22.13 & 19.75 & 17.94 & 16.00 & 13.81 & 10.54 &  8.18 &  7.18 \\
   50 &  17.31 & 14.01 & 12.63 & 11.48 & 10.28 &  9.00 &  7.00 &  5.54 &  4.94 \\
  100 &  12.19 & 10.04 &  9.02 &  8.21 &  7.38 &  6.47 &  5.04 &  3.99 &  3.56 \\
  200 &   8.68 &  7.11 &  6.40 &  5.83 &  5.22 &  4.57 &  3.59 &  2.86 &  2.56 \\ 
  500 &   5.61 &  4.54 &  4.08 &  3.70 &  3.33 &  2.91 &  2.28 &  1.82 &  1.63 \\
 1000 &   3.93 &  3.19 &  2.87 &  2.60 &  2.33 &  2.05 &  1.61 &  1.29 &  1.15 \\
 2000 &   2.71 &  2.24 &  2.02 &  1.84 &  1.66 &  1.46 &  1.14 &  0.91 &  0.82 \\
 5000 &   1.73 &  1.42 &  1.28 &  1.17 &  1.05 &  0.92 &  0.72 &  0.58 &  0.52 \\
10000 &   1.22 &  1.00 &  0.90 &  0.82 &  0.74 &  0.65 &  0.51 &  0.41 &  0.37 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

\begin{table}
\begin{center}
\caption{Critical values of test statistic $\tau_{\alpha, n, \kappa = 0.70}$ for
\textbf{empirical central normality} at $\kappa = 0.70$, as a function of significance 
level $\alpha$ and number of instances $n$. The values shown must be divided by $100$.}
\label{tab:empirical-central-normality-critical-statistic}
\begin{tabular}{l | r r r r r r r r r}

\toprule
$n$ \textbackslash $\alpha$ & 0.001 & 0.01 & 0.025 & 0.05 & 0.1 & 0.2 & 0.5 & 0.8 & 0.9 \\

\midrule
    5 & 159.27 & 66.78 & 57.06 & 50.55 & 43.42 & 34.89 & 22.28 & 14.92 & 12.10 \\
   10 &  41.82 & 30.82 & 27.01 & 24.31 & 21.58 & 18.53 & 13.82 & 10.23 &  8.75 \\
   20 &  29.44 & 23.65 & 21.23 & 19.16 & 17.01 & 14.67 & 11.10 &  8.42 &  7.35 \\
   50 &  18.95 & 15.56 & 13.96 & 12.74 & 11.35 &  9.82 &  7.51 &  5.83 &  5.16 \\ 
  100 &  13.49 & 11.00 &  9.91 &  8.98 &  8.03 &  7.04 &  5.43 &  4.25 &  3.76 \\
  200 &   9.65 &  7.99 &  7.22 &  6.64 &  5.96 &  5.21 &  4.05 &  3.17 &  2.81 \\
  500 &   6.37 &  5.32 &  4.82 &  4.44 &  4.03 &  3.57 &  2.79 &  2.18 &  1.92 \\
 1000 &   4.64 &  3.95 &  3.65 &  3.39 &  3.11 &  2.78 &  2.20 &  1.73 &  1.53 \\
 2000 &   3.61 &  3.18 &  2.96 &  2.78 &  2.57 &  2.32 &  1.89 &  1.50 &  1.33 \\
 5000 &   2.78 &  2.48 &  2.36 &  2.24 &  2.11 &  1.95 &  1.66 &  1.39 &  1.25 \\
10000 &   2.39 &  2.18 &  2.08 &  2.00 &  1.91 &  1.81 &  1.59 &  1.39 &  1.28 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

@@

{\centering \includegraphics[width=1\linewidth]{appendix_figure_2}

@@

\caption{Critical test statistic $\tau_{\alpha, n, \kappa = 0.80}$ of the (empirical) central normality test as function of $n$ for several values of significance level $\alpha$. The critical test statistics for central normality test are determined using fully normal data, whereas the statistics for the empirical variant are determined using centrally normal data, i.e. with fully normal data where 10\% of elements are replaced by outliers. emp: empirical}\label{fig:empirical-central-normality-critical-statistics}
\end{figure}

\section{Appendix D: Assessing transformations using simulated
data}\label{appendix-d}

\subsection{Ranking Box-Cox
transformations}\label{ranking-box-cox-transformations}

Box-Cox transformations were assessed in the same manner as Yeo-Johnson
transformations. The results are shown in Table
\ref{tab:comparison_methods_simulations_box_cox}.

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Box-Cox transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 10000 sequences each. The clean dataset consists of sequences derived through inverse Box-Cox transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
If necessary, each sequence was shifted so that every element had a strictly positive value. 
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_box_cox}
\begin{tabular}{l | l r r r r r r}

none                                  & &         7.65  &         6.96  &         7.28  &         6.58  &         7.45  &         6.78 \\
conventional                          & &         3.10  &         5.98  &         5.42  &         6.16  &         6.49  &         6.10 \\
conventional (z-score norm.)          & &         4.19  &         6.28  &         4.85  &         6.33  &         4.38  &         5.85 \\
conventional (robust scaling)         & &         4.20  &         6.28  &         4.78  &         6.33  &         4.30  &         5.86 \\
Raymaekers-Rousseeuw                  & &         4.46  &         3.55  &         5.39  &         3.69  &         6.29  &         5.84 \\ 
Raymaekers-Rousseeuw (z-score norm.)  & &         6.21  &         3.83  &         4.95  &         3.85  &         4.51  &         3.50 \\ 
Raymaekers-Rousseeuw (robust scaling) & &         6.12  &         3.82  &         4.97  &         3.85  &         4.52  &         3.50 \\
invariant                             & & \textbf{2.99} &         5.55  & \textbf{3.65} &         6.01  & \textbf{3.52} &         5.57 \\
robust invariant                      & &         6.09  & \textbf{2.75} &         3.71  & \textbf{2.20} &         3.55  & \textbf{2.00} \\

\subsection{Examples using clean data}\label{examples-using-clean-data}

Robust transformations are hypothesised to have a cost in efficiency for
data without outliers, i.e.~clean data. Here we draw nine sequences with
elements randomly drawn from a standard normal distribution
\(\mathcal{N}(0,1)\). Subsequently, we perform an inverse transformation
\(\left(\phi^{\lambda, 0, 1}\right)^{-1}\), with
\(\lambda \in \left(0, 2\right)\).

The sequences drawn resulted from the permutations of the number of
elements of each sequence (\(n \in \{30, 100, 500\}\)) and
transformation parameter for the inverse transformation
\(\lambda \in \{0.1, 1.0, 1.9\}\). Each sequence then underwent power
transformation using Box-Cox and Yeo-Johnson transformations. For
Box-Cox transformations, each sequence was shifted prior to inverse
transformation so that every element after inverse transformation would
be strictly positive.
